Certified Knowledge Compilation \\ with Application to Verified Model Counting

### 2023-03-04 doc/mc-paper.tex
Incremental progress

\newcommand{\indegree}{\textrm{indegree}}
\newcommand{\outdegree}{\textrm{outdegree}}

\newcommand{\simplify}[2]{#1|_{#2}}
\newcommand{\expand}{\mathit{Expand}}

@@

Assume that formula $\phi$ is 
represented as a set of clauses.
For partial assignment
$\passign$, the expression  $\simplify{\phi}{\passign}$ denotes the
reduced and simplified set of clauses obtained by 1) eliminating any
clause containing a literal $\lit$ such that $\passign(\lit) = 1$,
2) for the remaining clauses eliminating those literals $\lit$ for
which $\passign(\lit) = 0$, and 3) eliminating any duplicate clauses.
In doing these simplifications, we also track the {\em provenance}
of each simplified clause $C$, i.e., which of the original clauses simplified to become $C$.
More formally, for $C \in \simplify{\phi}{\passign}$, we let $\expand(C)$ denote
those clauses $C' \in \phi$, such that
$C' \subseteq C \cup \bigcup_{\lit \in \passign} \obar{\lit}$.
We then extend the definition of $\expand$ to any simplified formula
$\psi$ as $\expand(\psi) = \bigcup_{C \in \psi} \expand(C)$.

@@

The process for generating such a proof depends on the form of node $u$:
\begin{enumerate}
\item if $u$ is a literal $\lit$, then the formula
  $simplify{\phi}{\passign}$ must consist of the single unit clause
  $C = \lit$, such that $\expand{C}
  this clause.  That means that every other literal $\lit' \in C$ must
  have $\passign(\lit') = 0$, and therefore this input clause can
  serve as the target clause.
\item If $u$ is a disjunction node with children $u_1$ and $u_0$,
  then, since the node originated from a d-DNNF graph, there must be
  some variable $x$ such that either $u_1 = x$ or $u_1$ is a
  conjunction node containing $x$ as a child.  In either case, we
  recurse to node $u_1$ with the partial assignment $\passign \cup \{ x \}$.
  This will yield a sequence of proof steps
  $u \lor \obar{x} \lor \bigvee_{\lit \in \passign} \obar{l}$.
Similarly, either $u_0  = \obar{x}$ or $u_0$ is a conjunction node containing $\obar{x}$ as
  a child.  In either case, we recurse to node $u_0$
 with the partial assignment $\passign \cup \{ \obar{x} \}$.
  This will yield a sequence of proof steps
  $u \lor x \lor \bigvee_{\lit \in \passign} \obar{l}$.
  These recursive results can be combined, via two proof steps, with second and third defining clauses for $u$
  to generate the target clause for $u$.
\item If $u$ is a conjunction node, then we can divide its children
  into literal nodes and nonterminal nodes.
  \begin{enumerate}
    \item For each literal child
  $\lit$, we must prove that any total assignment $\alpha$, such that
  $\passign \subset \alpha$ must have $\alpha(\lit) = 1$.  In some
  cases, this can be done by simple Boolean constraint propagation.
  In other cases, we must prove that the formula
  $\simplify{\phi}{\passign \cap \{\obar{l}\}} is unsatisfiable.  We
  do so by writing the formula to a file, invoking a proof-generating
  SAT solver, and then importing the generated unsatisfiability prove
  as assertions in the CRAT file.
\item if there are multiple nonterminal children $u_1, u_2, \ldots,
  u_k$, it must be the case that the clauses in
  $\simplify{\phi}{\passign}$ can be partitioned into $k$ subsets
  $\psi_1, \psi_2, \ldots, \psi_k$ such that $\dependencyset(\psi_i)
  \cap \dependencyset(\psi_j) = \emptyset$ for $1 \leq i < j \leq k$,
  and we can match each node $u_i$ to subset $\psi_i$ based on its
  literals.  Moreover, each clause $C$ in each subset $\psi_i$ must be
  the simplified version of one or more input clauses $C'$ in the
  original formula $\phi$.  We define $\phi_i$ to consist of all such
  clauses $C'$ for each $C \in \psi_i$.

We do so using a union-find algorithm and match the variables in each partition
    $\phi_i$ to the POG node $u_i$ containing literals of the variables in
    $\dependencyset(\phi_i)$.
    We then recursively generate target clauses for nodes $u_1, u_2, \ldots, u_k$

\end{itemize}

\end{enumerate}

>>>>

\newcommand{\indegree}{\textrm{indegree}}
\newcommand{\outdegree}{\textrm{outdegree}}
\newcommand{\validate}{\textsf{validate}}

\newcommand{\simplify}[2]{#1|_{#2}}
\newcommand{\prov}{\mathit{Expand}}

@@

Assume that formula $\phi$ is 
represented as a set of clauses.
For partial assignment
$\passign$, the expression  $\simplify{\phi}{\passign}$ denotes the
reduced and simplified set of clauses obtained by 1) eliminating any
clause containing a literal $\lit$ such that $\passign(\lit) = 1$,
2) for the remaining clauses eliminating those literals $\lit$ for
which $\passign(\lit) = 0$, and 3) eliminating any duplicate clauses.
In doing these simplifications, we also track the {\em provenance}
of each simplified clause $C$, i.e., which of the (possibly multiple) original clauses simplified to become $C$.
More formally, for $C \in \simplify{\phi}{\passign}$, we let $\prov(C)$ denote
those clauses $C' \in \phi$, such that
$C' \subseteq C \cup \bigcup_{\lit \in \passign} \obar{\lit}$.
We then extend the definition of $\prov$ to any simplified formula
$\psi$ as $\prov(\psi) = \bigcup_{C \in \psi} \prov(C)$.

In the following we define a recursive routine $\validate(u, \passign,
\psi)$ taking as arguments 1) POG node $u$, 2) partial assignment
$\passign$, and 3) a set of clauses $\psi \subsetq \phi$.  The outcome of its execution should be to add a number of assertions culminating with
the addition of the {\em target clause}
$u \lor \bigvee_{\lit \in \passign} \obar{\lit}$, indicating that any total
assignment $\assign$ such that $\passign \subseteq \assign$
will, when extended to extension variables, will assign $\assign(u) = 1$.
The top-level call has $u = r$, $\passign = \emptyset$, and $\psi = \phi$.  The result will therefore be to add the unit clause $r$ to the proof.

@@

The process for generating such a proof depends on the form of node $u$:
\begin{enumerate}
\item if $u$ is a literal $\lit$, then the formula
  $simplify{\phi}{\passign}$ must consist of the single unit clause
  $C = \lit$, such that any $C' \in \prov{C}$ must have $C' \subseteq \{ \lit \} \bigcup_{\lit \in \passign} \obar{\lit}$.
  Any of these can
  serve as the target clause.
\item If $u$ is a disjunction node with children $u_1$ and $u_0$,
  then, since the node originated from a d-DNNF graph, there must be
  some variable $x$ such that either $u_1 = x$ or $u_1$ is a
  conjunction node containing $x$ as a child.  In either case, we
  recurse to node $u_1$ with the partial assignment $\passign \cup \{ x \}$.
  This will yield a sequence of proof steps
  $u \lor \obar{x} \lor \bigvee_{\lit \in \passign} \obar{\lit}$.
Similarly, either $u_0  = \obar{x}$ or $u_0$ is a conjunction node containing $\obar{x}$ as
  a child.  In either case, we recurse to node $u_0$
 with the partial assignment $\passign \cup \{ \obar{x} \}$.
  This will yield a sequence of proof steps
  $u \lor x \lor \bigvee_{\lit \in \passign} \obar{\lit}$.
  These recursive results can be combined, via two proof steps, with second and third defining clauses for $u$
  to generate the target clause for $u$.
\item If $u$ is a conjunction node, then we can divide its children
  into a set of literal nodes $L$ and a set of nonterminal nodes $u_1, u_2, \ldots, u_k$.
  \begin{enumerate}
    \item For each literal 
  $\lit \in L$, we must prove that any total assignment $\alpha$, such that
  $\passign \subset \alpha$ must have $\alpha(\lit) = 1$.  In some
  cases, this can be done by simple Boolean constraint propagation (BCP).
  In other cases, we must prove that the formula
  $\simplify{\phi}{\passign \cap \{\obar{\lit}\}}$ is unsatisfiable.  We
  do so by writing the formula to a file, invoking a proof-generating
  SAT solver, and then importing the generated unsatisfiability prove
  as assertions in the CRAT file.
\item If there is a single nonterminal child ($k = 1$), we can recurse on node $u_1$
  with partial assignment $\passign \cup \bigcup_{\lit \in L} \lit$.
\item If there are multiple nonterminal children ($k > 1$)
  it must be the case that the clauses in
  $\simplify{\phi}{\passign}$ can be partitioned into $k$ subsets
  $\psi_1, \psi_2, \ldots, \psi_k$ such that $\dependencyset(\psi_i)
  \cap \dependencyset(\psi_j) = \emptyset$ for $1 \leq i < j \leq k$,
  and we can match each node $u_i$ to subset $\psi_i$ based on its
  literals.
  For each $i$ such that $1 \leq i \leq k$, we can then recurse on node $u_i$
  with partial assignment $\passign \cup \bigcup_{\lit \in L} \lit$, and based on the input clauses in $\prov(\psi_i)$.
\end{enumerate}
  We then combine as hints the results from the BCP and SAT calls for
  the literals, the recursively computed target clauses, and all but
  the first defining clause for the conjunction operation to generate the target clause for node $u$.
\end{enumerate}
